Supplementary information: In-silico molecular enrichment and clearance of the human intracranial space

### 2025-02-27 paper/supplementary_main.tex
Split supplementary into separate file with independent references...

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(\Cref{fig:results1}A, \Cref{fig:csf}A--C). The coordinates in these
3D domains is denoted by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
towards the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

@@

In this section, we describe the mathematical
model~\eqref{eq:multi_transport}, the associated definitions,
interface and boundary conditions in further detail. We refer to the
main text for explicit parameter values while giving the general,
abstract form here, and to the reference~\cite{masri2024modelling, laurino2019derivation} for
the derivation and analysis of this 3D-1D model. Recall that we model
a concentration field $c = c(x, t)$ for $x \in \Omega$ and $t > 0$
defined in $\Omega_{\rm PAR}$ and $\Omega_{\rm CSF}$ separately. In
each of these domains, $c$ satisfies
%\mer{MER: Rami asks if it should be a $\phi$ in the advective term here. We can include it, or drop it, but in the latter case, perhaps we should clarify that u is an average/superficial/Darcy velocity.}
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma
  %= f
   = 0
  \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction, and thus $c$ here generally represents the intrinsic (in contrast to the superficial) concentration~\cite{sykova2008diffusion}. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivascular networks $\Lambda_a, \Lambda_v$. To summarize,
\begin{equation*}
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\end{equation*}
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$ if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$, in which case the interaction with $\Omega_{\rm{CSF}}$ is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

@@

fields $R_c$ and $R_r$ (\Cref{fig:csf}G, I), associated with the

@@

  CSF space in the respective cycle (\Cref{fig:csf}F, H). More

@@

  1.5$mm, and CSF viscosity $\mu$ given in~\Cref{tab:parameters}.  For

@@

We use the theoretical framework previously introduced by Gjerde et al.~\cite{gjerde2023directional} to compute an analytic estimate of the time-averaged (or \emph{net}) flow rates $\langle Q'_i \rangle$ (mm$^3$/s) induced by peristaltic pumping in a perivascular network $\Lambda = \cup_i \Lambda_i$. The motion of the (inner) vascular wall is described by a periodic traveling (peristaltic) wave of relative amplitude $\varepsilon$, wave length $\lambda$ (mm) and frequency $f$, acting normal to the wall. By definition, the wave number is $k = 2 \pi/\lambda$ and the angular frequency is $\omega = 2 \pi f$. Each PVS segment $\Lambda_i$ has length $L_i$ with wave-relative length $\ell_i = k L_i$, baseline inner radius $R_1^i$, fixed outer radius $R_2^i$, and outer-to-inner ratio $\beta_i = R_2^i/R_1^i$. These geometric parameters and the assumption of annular cylindrical PVS segments yield hydraulic resistances $\mathcal{R}_{o, i}$ and additional characteristic parameters, see \cite{gjerde2023directional} for the complete definitions and schematics. Since the analytical estimate is derived under the assumption that $k L_i \approx \mathcal{O} (1)$~\cite{gjerde2023directional} and has been verified against numerical simulations for $k L_i$ of the order $10^{-1}-10^2$~\cite[Table I]{gjerde2023directional}, we consider it applicable for the wave lengths and vascular network data considered here in which $k L_i$ range from 0.15 to 30 for the strong vasomotion and 0.0015 to 0.30 for the cardiac waves (\Cref{tab:pvs:parameters}).
\begin{table}
\centering
  \small
  \begin{tabular}{llrc}
    \toprule
    Parameter & Description & Value(s)  & Ref.\\ 
    \midrule
    Relative PVS size $\beta$ & $\beta = \beta_i = R_2^i / R_1^i$ & 2 & $\approx$\cite{eide2024functional} \\
    Wave frequency $f$ & Traveling wave frequency of vascular motion & $0.1-1.0$ Hz & \discuss{\cite{gjerde2023directional}} \\
    Wave length $\lambda$ & Traveling wave length of vascular motion & $0.02-2$ mm & \discuss{\cite{broggini2024long, gjerde2023directional}} \\
    Wave amplitude & Relative amplitude of inner wall motion & $1-10\%$ & \discuss{\cite{gjerde2023directional}} \\
    \midrule
    Cardiac frequency $f_c$ & Frequency of human cardiac pulse wave  & $1.0$ Hz & \cite{jung2021novel} \\
    Cardiac wave length $\lambda_c$ & Wave length of human cardiac pulse wave & $2.0$m & \\
    Cardiac amplitude $\varepsilon_c$ & Wall displacement amplitude of cardiac pulse wave & 1\%  & \\
    \midrule
    Vasomotion frequency $f_v$ & Frequency of slow vasomotion wave  & $0.1$ Hz & \cite{jung2021novel} \\
    Vasomotion wave length $\lambda_v$ & Wave length of vasomotion wave & $0.02$ m & \cite{broggini2024long, gjerde2023directional} \\
    Vasomotion amplitude $\varepsilon_v$ & Wall displacement amplitude of vasomotion wave & 10\%  & \cite{broggini2024long, gjerde2023directional}\\
    %Vascular radii $\{r_{o_i}\}$ & Vascular network radii data (avg $\pm$ std, $\min$, $\max$) & $X$, $X$, $X$  mm & \cite{hodneland2019new} \\
    %Vascular branch lengths $\{L_{i}\}$ & Path length between vasc. junctions (avg $\pm$ std, $\min$, $\max$) & $X \pm Y$, $X$, $X$  mm & \discuss{--} \\
    \bottomrule
  \end{tabular}
  \caption{Perivascular flow induced by vascular wall motion: overview of parameters.}
  \label{tab:pvs:parameters}
\end{table}
% Vascular radii r_o (min, max, avg, std) (mm):  ....
% Vascular branch lengths L (avg, std, min, max) (mm): ...

@@

This theoretical formalism~\cite{gjerde2023directional} is defined
relative to a network in the form of a directed, bifurcating tree with
a single supply node/root $i_0$. To extend to a network of cerebral
arteries with multiple supply nodes (such as the basilar and two
internal carotid arteries in the current data
set~\cite{hodneland2019new}), we separate the network $\Lambda$ into
edge-disjoint subnetworks $\Lambda^1, \Lambda^2, \Lambda^3$, one for
each of the supply nodes (\Cref{fig:results1}A--B,
\Cref{fig:app:peristalsis}). Each node is assigned to the subnetwork
associated with the nearest supply node, and edges between nodes are
preserved (\Cref{fig:app:peristalsis}B). Next, we compute a minimal,
bifurcating and directed tree representation of each subnetwork:
$\mathcal{T}^1, \mathcal{T}^2, \mathcal{T}^3$
(\Cref{fig:app:peristalsis}C). Each tree $\mathcal{T}^j = \cup E_n^j$
consists of the subset of the nodes from $\Lambda^j$ that have degree
1 (are leaf or root nodes) or degree $3$ (are true bifurcation
points), and each path between nodes with degree $\not = 2$ in
$\Lambda^j$ is represented in $\mathcal{T}^j$ by an edge $E^j$ with
edge length $L$ corresponding to the total length of the original path
and edge radius $R^j$ as the average of the path radii.

>>>>

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(Fig.~1A, Fig.~2A--C). The coordinates in these
3D domains is denoted by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
towards the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

@@

In this section, we describe the mathematical model (eq.~(1)), the
associated definitions, interface and boundary conditions in further
detail. We refer to the main text for explicit parameter values while
giving the general, abstract form here, and to the
reference~\cite{masri2024modelling, laurino2019derivation} for the
derivation and analysis of this 3D-1D model. Recall that we model a
concentration field $c = c(x, t)$ for $x \in \Omega$ and $t > 0$
defined in $\Omega_{\rm PAR}$ and $\Omega_{\rm CSF}$ separately. In
each of these domains, $c$ satisfies
%\mer{MER: Rami asks if it should be a $\phi$ in the advective term here. We can include it, or drop it, but in the latter case, perhaps we should clarify that u is an average/superficial/Darcy velocity.}
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma
  %= f
   = 0
  \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction, and thus $c$ here generally represents the intrinsic (in contrast to the superficial) concentration~\cite{sykova2008diffusion}. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivascular networks $\Lambda_a, \Lambda_v$. To summarize,
\begin{equation*}
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\end{equation*}
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$ if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$, in which case the interaction with $\Omega_{\rm{CSF}}$ is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

@@

fields $R_c$ and $R_r$ (Fig.~2G, I), associated with the

@@

  CSF space in the respective cycle (Fig.~2F, H). More

@@

  1.5$mm, and CSF viscosity $\mu$ given in~Table.~1.  For

@@

We use the theoretical framework previously introduced by Gjerde et al.~\cite{gjerde2023directional} to compute an analytic estimate of the time-averaged (or \emph{net}) flow rates $\langle Q'_i \rangle$ (mm$^3$/s) induced by peristaltic pumping in a perivascular network $\Lambda = \cup_i \Lambda_i$. The motion of the (inner) vascular wall is described by a periodic traveling (peristaltic) wave of relative amplitude $\varepsilon$, wave length $\lambda$ (mm) and frequency $f$, acting normal to the wall. By definition, the wave number is $k = 2 \pi/\lambda$ and the angular frequency is $\omega = 2 \pi f$. Each PVS segment $\Lambda_i$ has length $L_i$ with wave-relative length $\ell_i = k L_i$, baseline inner radius $R_1^i$, fixed outer radius $R_2^i$, and outer-to-inner ratio $\beta_i = R_2^i/R_1^i$. These geometric parameters and the assumption of annular cylindrical PVS segments yield hydraulic resistances $\mathcal{R}_{o, i}$ and additional characteristic parameters, see \cite{gjerde2023directional} for the complete definitions and schematics. Since the analytical estimate is derived under the assumption that $k L_i \approx \mathcal{O} (1)$~\cite{gjerde2023directional} and has been verified against numerical simulations for $k L_i$ of the order $10^{-1}-10^2$~\cite[Table I]{gjerde2023directional}, we consider it applicable for the wave lengths and vascular network data considered here in which $k L_i$ range from 0.15 to 30 for the strong vasomotion and 0.0015 to 0.30 for the cardiac waves (\Cref{tab:pvs:parameters}).
\begin{table}
\centering
  \small
  \begin{tabular}{llrc}
    \toprule
    Parameter & Description & Value(s)  & Ref.\\ 
    \midrule
    Relative PVS size $\beta$ & $\beta = \beta_i = R_2^i / R_1^i$ & 2 & $\approx$\cite{eide2024functional} \\
    Wave frequency $f$ & Traveling wave frequency of vascular motion & $0.1-1.0$ Hz & \cite{gjerde2023directional} \\
    Wave length $\lambda$ & Traveling wave length of vascular motion & $0.02-2$ mm & \cite{broggini2024long, gjerde2023directional} \\
    Wave amplitude & Relative amplitude of inner wall motion & $1-10\%$ & \cite{gjerde2023directional} \\
    \midrule
    Cardiac frequency $f_c$ & Frequency of human cardiac pulse wave  & $1.0$ Hz & \cite{jung2021novel} \\
    Cardiac wave length $\lambda_c$ & Wave length of human cardiac pulse wave & $2.0$m & \\
    Cardiac amplitude $\varepsilon_c$ & Wall displacement amplitude of cardiac pulse wave & 1\%  & \\
    \midrule
    Vasomotion frequency $f_v$ & Frequency of slow vasomotion wave  & $0.1$ Hz & \cite{jung2021novel} \\
    Vasomotion wave length $\lambda_v$ & Wave length of vasomotion wave & $0.02$ m & \cite{broggini2024long, gjerde2023directional} \\
    Vasomotion amplitude $\varepsilon_v$ & Wall displacement amplitude of vasomotion wave & 10\%  & \cite{broggini2024long, gjerde2023directional}\\
    \bottomrule
  \end{tabular}
  \caption{Perivascular flow induced by vascular wall motion: overview of parameters.}
  \label{tab:pvs:parameters}
\end{table}
% Vascular radii r_o (min, max, avg, std) (mm):  ....
% Vascular branch lengths L (avg, std, min, max) (mm): ...

@@

This theoretical formalism~\cite{gjerde2023directional} is defined
relative to a network in the form of a directed, bifurcating tree with
a single supply node/root $i_0$. To extend to a network of cerebral
arteries with multiple supply nodes (such as the basilar and two
internal carotid arteries in the current data
set~\cite{hodneland2019new}), we separate the network $\Lambda$ into
edge-disjoint subnetworks $\Lambda^1, \Lambda^2, \Lambda^3$, one for
each of the supply nodes (Fig.~1A--B,
\Cref{fig:app:peristalsis}). Each node is assigned to the subnetwork
associated with the nearest supply node, and edges between nodes are
preserved (\Cref{fig:app:peristalsis}B). Next, we compute a minimal,
bifurcating and directed tree representation of each subnetwork:
$\mathcal{T}^1, \mathcal{T}^2, \mathcal{T}^3$
(\Cref{fig:app:peristalsis}C). Each tree $\mathcal{T}^j = \cup E_n^j$
consists of the subset of the nodes from $\Lambda^j$ that have degree
1 (are leaf or root nodes) or degree $3$ (are true bifurcation
points), and each path between nodes with degree $\not = 2$ in
$\Lambda^j$ is represented in $\mathcal{T}^j$ by an edge $E^j$ with
edge length $L$ corresponding to the total length of the original path
and edge radius $R^j$ as the average of the path radii.
